---
title: "Data cleaning (AI news database)"
subtitle: "Trabajo Fin de Master"
author: "Isabel Molero López"
date: "Sys.Date()"
output: html_notebook
  toc: true
---

# 1. Introduction

This file is the second part of a series of rmds that I am doing for my Master's 
Thesis. In the previous one ("web-scraping AI articles) I extracted the content 
of newspaper articles in a csv. Now I am going to read the csv and clean the data
to be able to perform visualizations and text analysis.

# 2. Load libraries

```{r message=FALSE}
library(tidyverse)
library(lubridate)
```


# 3. Import data

```{r message=FALSE}
ai_eldiario_row <- read_csv("../data/eldiario_news2.csv") %>%
  filter(!duplicated(link))

ai_elmundo_row <- read_csv("../data/elmundo_news2.csv") %>%
  filter(!duplicated(link)) %>% 
  filter(!duplicated(title)) %>% 
  mutate(link = str_replace(link, "//www.elmundo.es/economia/innovadores/2019/02/19/5c6ae9b721efa0fb378b46d4.html", "https://www.elmundo.es/economia/innovadores/2019/02/19/5c6ae9b721efa0fb378b46d4.html")) %>% 
  mutate(link = str_replace(link, "^//", "https://")) %>% 
  filter(!is.na(title))

text_eldiario_row <- read_csv("../data/eldiario_text2.csv") %>%
  filter(!duplicated(link))

text_elmundo_row <- read_csv("../data/elmundo_text2.csv") %>%
  filter(!duplicated(link))
```

# 4. Merge dataframes

```{r}
eldiario_all_row <- ai_eldiario_row %>% 
  full_join(text_eldiario_row, by = "link")

elmundo_all_row <- ai_elmundo_row %>% 
  full_join(text_elmundo_row, by = "link")
```

Now I am going to remove the dataframe that I am not going to use

```{r}
rm(st = ai_eldiario_row, ai_elmundo_row, text_eldiario_row, text_elmundo_row)
```


# 5. Data cleaning

## Date

```{r}
eldiario_all <- eldiario_all_row %>%
  mutate(date = str_split(date, " - ", simplify = TRUE)[, 1]) %>%
  separate(date, into = c("day", "month", "year"), sep = " de ") %>%
  mutate(date = dmy(paste(day, month, year)),
         date = ymd(date)) %>%
  select(-day, -month, -year)
```

```{r}
elmundo_all <- elmundo_all_row %>%
  mutate(date = str_sub(date, start = 1, end = 10),
         date = ymd(date))

elmundo_all <- elmundo_all_row %>% 
  mutate(date = ymd_hms(date, tz = "UTC"),
         date = ymd(format(date, "%Y-%m-%d")))
```

